Is Archaeology a science? Insights and imperatives from 10,000 articles and a year of reproducibility reviews

### 2025-04-19 analysis/paper/paper.qmd
onto comments from R2

In their paper celebrating the 40th anniversary of this journal Torrence et al [-@torrenceFortyYearsStill2015] noted that reproducibility was an issue important to the reputation and sustainability of the discipline, and necessary for archaeological science to behave like a science. As part of the celebration of the 50th anniversary, and of Torrence's leadership of the journal, my contribution revisits these topics of archaeology's status as a science, this journal's place in the landscape of archaeological science, and how the journal has responded to a growing recognition of the importance of reproducibility. I first present bibliometric evidence of the position of archaeology as a whole, and this journal in particular, in the sciences. Next I report on the journal's progress in supporting reproducible research, and my work doing a new kind of peer review for JAS, one that evaluates the computational reproducibilty of the research submitted for publication. Finally, I analyse twelve months of reproducibility reviews to identify common weaknesses in the ways archaeologists are working currently, and provide simple recommendations for researchers to overcome these and contribute to the improvement computational reproducibility in archaeological science.

@@

source("analysis/code/008-redraw-Fanelli-and-Glanzel-Fig-2.R")

@@

# Number of authors ------------------
boxlplot_n_authors <- 
  ggplot() +
  # boxplot of data from this study
  geom_boxplot(data = items_df %>% 
  filter(!is.na(year)),
  aes(1, log(authors_n)),
               size = 1)  +
  # boxplot of data from Fanelli & Glänzel Fig 2
  geom_boxplot(data = sim_data %>% 
  filter(Variable == "N. of authors (ln)",
         Category %in% c("h", "p", "s")),
  aes(1, Value, 
      group = Category),
  size = 1,
  colour = color, 
  alpha = 0)  +
  # annotations from Fanelli & Glänzel Fig 2
  geom_text_repel(data = sim_data %>%  
    filter(Variable == "N. of authors (ln)", 
           Category %in% subcat_order) %>%  
    group_by(Category) %>%  
    summarise(y = median(Value)) %>%
    mutate(
      label = as.character(Category)
    ),
                  aes(c(0.75, 1, 1.25), y, label = label),
                  color = color,
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
  scale_y_continuous(limits = c(0, 5)) +
  scale_x_continuous(labels = NULL) +
  theme_minimal(base_size = base_size) + 
  theme(panel.grid  = element_blank()) +
  ylab("N. of authors (ln)") +
  xlab("Collaborator group size") 
```

@@

boxlplot_rel_title_length <- 
items_df_title %>% 
  filter(!is.na(year)) %>% 
  ggplot(aes(1,
             relative_title_length)) +
  geom_boxplot(
               size = 1)  +
  # boxplot of data from Fanelli & Glänzel Fig 2
  geom_boxplot(data = sim_data %>% 
  filter(Variable == "Relative title length (ln)",
         Category %in% c("h", "p", "s")),
  aes(1, Value, 
      group = Category),
  size = 1,
  colour = color, 
  alpha = 0)  +
  # annotations from Fanelli & Glänzel Fig 2
  geom_text_repel(data = sim_data %>%  
    filter(Variable == "Relative title length (ln)", 
           Category %in% subcat_order) %>%  
    group_by(Category) %>%  
    summarise(y = median(Value)) %>%
    mutate(
      label = as.character(Category)
    ),
                  aes(c(0.75, 1, 1.25), y, label = label),
                  bg.colour = "white", 
    colour = color,
                  bg.r = .2, 
                  force = 0) +
  scale_y_continuous(limits = c(-4.5, 3),
                     breaks =  seq(-5, 5, 1),
                     labels = seq(-5, 5, 1)) +
  scale_x_continuous(labels = NULL) +
  theme_minimal(base_size = base_size) +
  theme(panel.grid  = element_blank()) +
  ylab("Ratio of title length to article length (ln)") +
  xlab("Relative title length")

# Number of pages ------------------
boxlplot_n_pages <- 
items_df %>% 
  ggplot(aes(1,
             log(pages_n))) +
  geom_boxplot(
               size = 1)  +
  # boxplot of data from Fanelli & Glänzel Fig 2
  geom_boxplot(data = sim_data %>% 
  filter(Variable == "N. of pages (ln)",
         Category %in% c("h", "p", "s")),
  aes(1, Value, 
      group = Category),
  size = 1,
  colour = color, 
  alpha = 0)  +
  # annotations from Fanelli & Glänzel Fig 2
  geom_text_repel(data = sim_data %>%  
    filter(Variable == "N. of pages (ln)", 
           Category %in% subcat_order) %>%  
    group_by(Category) %>%  
    summarise(y = median(Value)) %>%
    mutate(
      label = as.character(Category)
    ),
                  aes(c(0.75, 1, 1.25), y, label = label),
                  bg.colour = "white",
    colour = color, 
                  bg.r = .2, 
                  force = 0) +
  scale_y_reverse(limits = c(5, 0)) +
  scale_x_continuous(labels = NULL) +
  theme_minimal(base_size = base_size) + 
  theme(panel.grid  = element_blank()) +
  ylab("N. of pages (ln)") +
  xlab("Article length")

@@

# plot
boxlplot_price_index <- 
items_df %>% 
  ggplot(aes(1,
             prices_index)) +
  geom_boxplot(
               size = 1)  +
  # boxplot of data from Fanelli & Glänzel Fig 2
  geom_boxplot(data = sim_data %>% 
  filter(Variable == "Price's index",
         Category %in% c("h", "p", "s")),
  aes(1, Value, 
      group = Category),
  size = 1,
  colour = color,  
  alpha = 0)  +
  # annotations from Fanelli & Glänzel Fig 2
  geom_text_repel(data = sim_data %>%  
    filter(Variable == "Price's index", 
           Category %in% subcat_order) %>%  
    group_by(Category) %>%  
    summarise(y = median(Value)) %>%
    mutate(
      label = as.character(Category)
    ),
                  aes(c(0.75, 1, 1.25), y, label = label),
                  bg.colour = "white", colour = color, 
                  bg.r = .2, 
                  force = 0) +
  scale_y_continuous(limits = c(0, 1)) +
  scale_x_continuous(labels = NULL) +
  theme_minimal(base_size = base_size) +
  theme(panel.grid  = element_blank()) +
  ylab("Prop. refs in last 5 years") +
  xlab("Recency of references")

@@

# plot
boxlplot_shannon_index <- 
shannon_per_item %>% 
  filter(!is.na(year)) %>% 
  ggplot(aes(1,
             shannon)) +
  geom_boxplot(
               size = 1)  +
   # boxplot of data from Fanelli & Glänzel Fig 2
  geom_boxplot(data = sim_data %>% 
  filter(Variable == "Shannon div. of sources",
         Category %in% c("h", "p", "s")),
  aes(1, Value, 
      group = Category),
  size = 1,
  colour = color,  
  alpha = 0)  +
  # annotations from Fanelli & Glänzel Fig 2
  geom_text_repel(data = sim_data %>%  
    filter(Variable == "Shannon div. of sources", 
           Category %in% subcat_order) %>%  
    group_by(Category) %>%  
    summarise(y = median(Value)) %>%
    mutate(
      label = as.character(Category)
    ),
                  aes(c(0.75, 1, 1.25), y, label = label),
    colour = color, 
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
  scale_y_reverse(limits = c(6, 0)) +
  scale_x_continuous(labels = NULL) +
  theme_minimal(base_size = base_size)  +
  theme(panel.grid  = element_blank()) +
  ylab("Shannon Index") +
  xlab("Diversity of references")
```

```{r}
#| label: fig-compare-other-fields
#| fig-cap: Distributions of article characteristics hypothesised to reflect the level of consensus. The boxplot shows the distribution of values of archaeology articles. The thick black line in the middle of the boxplot is the median value, the box represents the inter-quartile range (the range between the 25th and 75th percentiles, where 50% of the data are located), and individual points represent outliers. The thin grey horizontal lines in each boxplot indicate the median values computed by Fanelli and Glanzel (2013), where p = physics, s = social sciences, h = humanities. ln denotes the natural logarithm, or logarithm to the base e. 
#| fig-height: 4

@@

@fig-compare-other-fields shows the distribution of bibliometric variables for archaeology in context of data from other fields presented by @fanelliBibliometricEvidenceHierarchy2013. The most striking indicator of archaeology as a hard science is the number of authors, where it is between the social sciences and physics. Archaeology is a close fit with the social sciences in relative title length. It is between the the social sciences and humanities in recency of references and diversity of references. The clearest indicator of archaeology as a soft science is article length where it is similar to the humanities. Overall archaeology does not sit squarely at either end of the hard-soft spectrum. It is generally not a harder science than the social sciences, with the exception of collaborator group sizes.

@@

The most striking change over time is in the increase in the number of authors, which has the highest r-squared value of these metrics. One interesting detail evident in @fig-change-over-time is the increase in the range of diversity of references after about 2010. This may be due to some broader changes in academic publishing around this time, such as moves to digital-first continuous publishing, new journals appearing (e.g. *Archaeological and Anthropological Sciences* in 2009 and *Journal of Island & Coastal Archaeology* in 2010), and non-archaeology journals becoming more relevant to archaeologists. For example, *PLOS ONE* received its first impact factor in 2010 and in 2011 Nature's *Scientific Reports* began publishing [@malashichev2017open].

@@

The PCA results in panel G show of @fig-variation-by-journal that PC1 captures most of the variance in the metrics (`r variance_explained[1]`%) and is reasonable proxy for the hard-soft spectrum, with *Journal of Cultural Heritage* representing the hard extreme on the right and *Journal of Archaeological Research* representing the soft extreme on the left. The distribution of PC1 values is skewed left, with most of the journals concentrated at the harder end of the spectrum. The distribution of PC2 values reveals additional structure to the data and can be roughly separated into generalist journals in the negative range (e.g. *American Antiquity*, *Antiquity*, *Advances in Archaeological Practice*), and more specialised journals in the positive range (e.g. *Environmental Archaeology*, *Geoarchaeology*, *Archaeologial Research in Asia*, *Journal of Island and Coastal Archaeology*). The *Journal of Archaeological Science* sits about midway between these two groups, reflecting its relevance to specialised communities of practice in archaeology, as well as general

@@

With many disciplines making increasing use of computationally intensive analyses in recent years there has been renewed interest in reproducibility [@levequeReproducibleResearchScientific2012]. In part, this is because computationally intensive research is difficult to communicate within the constraints of the methods section of a traditional journal article — the reader needs also needs the computer code written by the original authors, not just the article text. There is also the broader context of rising pressure to publish in prestigious journals and intense competition for funds that create strong incentives for malpractice in research [@edwards2017academic]. These two factors have led to widespread concerns of a reproducibility crisis in many fields [@Baker2016]. Estimates of scientific reproducibility in several fields confirm the extent of this problem. Empirical replications of 100 studies published in three psychology journals found that 36% of replications had statistically significant results, compared to 96% of the original studies [@OpenScienceCollaboration2015]. Similar empirical replications of large numbers of social science studies and experiemental economics studies successfully replicated 61% and 62% of their target studies respectively [@camererEvaluatingReplicabilitySocial2018; @camererEvaluatingReplicabilityLaboratory2016].

@@

base_size <- 4

@@

```{r}
#| label: fig-aer-summary
#| fig-cap: "Summary of reproducibility reviews for JAS. A: Primary software used for the computational analysis reported in a manuscript. B: Computational or statistical method used by the authors (GMM = geometric morphometrics, Frequentist = hypothesis tests such as chi-square and ANOVA, AI/ML = artifical intellegence and machine learning, including neural networks and deep learning, Network = statitical analysis of social networks, 3D = analysis of 3D data such as artefact models, Composition = compositional analysis of artefacts). C: Locations where authors deposited their code and data files. D: Issues that prevented the reproducible review from succeeding on the first attempt. E: Relationship between software used and issues that make research irreproducible."
#| fig-height: 3

@@

At the time of writing (January 2025) we have completed 47 reproducibility reviews of 25 manuscripts submitted to JAS. Of these, 11 have been published in JAS to date. Seven of these eleven papers fully passed the reproducibility review, resulting in a success rate, by one measure, of 63%. Four of the seven papers could be fully reproduced on my first attempt, the others required additional input from the authors. For comparison with reproducibility studies in other fields reported above, the seven fully reproducible papers should be divided by the 25 reviewed for reproducibility, resulting in a 28% success rate. Expanding the denominator to include the total number of research articles published in JAS from May 2024 (when the first article to pass the reproducibility review, @herskindComputationalLinguisticMethodology2024, was published) to January 2025 (n = 97), which are articles that could have been eligible for reproducibility review had the authors used an open source programming language (e.g. instead of commercial software such as Microsoft Excel or SPSS, etc.) the success rate is 7%. In any case, the computational reproducibility of archaeological research is generally on the low end of the distribution of values available from a variety of hard and soft sciences.

@@

Third, authors should include in their compendium a README document that describes to readers the folders and files contained in the project. The README file is typically the first file that a reader will look at in a compendium so it should include brief instructions to guide the user to a successful reproduction of the original results (e.g. what order to run the code files in). A README should also briefly describe the contents of the compendium, where other necessary files can be obtained (e.g. data files not included in the compendium), and the key software packages needed and the version numbers that the authors used.

@@

Computational reproducibility is not a panacea, results that are fully reproducible can contain errors and fraud. It is no guarantee of code quality, or that statistics have been used appropriately (cf. Vaiglova 2025; Crema 2025), or that data management is consistent with FAIR and CARE principles [@carrollOperationalizingCAREFAIR2021]. It is also time-consuming for authors to ensure their computational work can be reproduced, and for reviewers to evaluate. In an professional environment where job security and career progression is often associated with pressure to publish many high-impact papers, demands for authors to spend time on reproducibility, resulting in less time for publishing more papers, may seem frustrating. This may seem especially unfair to early career researchers on short-term contracts, who may feel the goalposts for career success are being moved and that they are being asked to do more work their graduate training has not prepared them for. This highlights the need for a culture shift among senior archaeological scientists to value reproducibility in hiring and promotion decision-making. Professors must contribute to this shift by nurturing a culture of reanalysis and reproducibility in their teaching, for example by using replication assignments and by training students in the most current best practices and tools for reproducible research, such as R and Python [@marwickHowUseReplication2020].

>>>>

In their paper celebrating the 40th anniversary of this journal Torrence et al [-@torrenceFortyYearsStill2015] noted that reproducibility was an issue important to the reputation and sustainability of the discipline, and necessary for archaeological science to behave like a science. As part of the celebration of the 50th anniversary, and of Torrence's leadership of the journal, my contribution revisits these topics of archaeology's status as a science, this journal's place in the landscape of archaeological science, and how the journal has responded to a growing recognition of the importance of reproducibility. I first present bibliometric evidence of the position of archaeology as a whole, and this journal in particular, in the sciences. Next I report on the journal's progress in supporting reproducible research, and my work doing a new kind of peer review for JAS, one that evaluates the computational reproducibilty of the research submitted for publication. Finally, I analyse twelve months of reproducibility reviews to identify common weaknesses in the ways archaeologists are working currently, and provide simple recommendations for researchers to overcome these and contribute to the improvement of computational reproducibility in archaeological science.

@@

source(here::here("analysis/code/008-redraw-Fanelli-and-Glanzel-Fig-2.R"))

@@

# Number of authors ------------------
boxlplot_n_authors <- 
  ggplot() +
  # boxplot of data from this study
  geom_boxplot(data = items_df %>% 
  filter(!is.na(year)),
  aes(1, log(authors_n)),
               size = 1)  +
  # boxplot of data from Fanelli & Glänzel Fig 2
  geom_boxplot(data = sim_data %>% 
  filter(Variable == "N. of authors (ln)",
         Category %in% c("h", "p", "s")),
  aes(1, Value, 
      group = Category),
  size = 1,
  colour = color, 
  alpha = 0)  +
  # annotations from Fanelli & Glänzel Fig 2
  geom_text_repel(data = sim_data %>%  
    filter(Variable == "N. of authors (ln)",
           Category %in% c("h", "p", "s")) %>%  
    group_by(Category) %>%  
    summarise(y = median(Value)) %>%
    mutate(
      label = as.character(Category)
    ),
                  aes(c(0.75, 1, 1.25), y, label = label),
                  color = color,
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
  scale_y_continuous(limits = c(0, 5)) +
  scale_x_continuous(labels = NULL) +
  theme_minimal(base_size = base_size) + 
  theme(panel.grid  = element_blank()) +
  ylab("N. of authors (ln)") +
  xlab("Collaborator group size") 
```

@@

boxlplot_rel_title_length <- 
items_df_title %>% 
  filter(!is.na(year)) %>% 
  ggplot(aes(1,
             relative_title_length)) +
  geom_boxplot(
               size = 1)  +
  # boxplot of data from Fanelli & Glänzel Fig 2
  geom_boxplot(data = sim_data %>% 
  filter(Variable == "Relative title length (ln)",
         Category %in% c("h", "p", "s")),
  aes(1, Value, 
      group = Category),
  size = 1,
  colour = color, 
  alpha = 0)  +
  # annotations from Fanelli & Glänzel Fig 2
  geom_text_repel(data = sim_data %>%  
    filter(Variable == "Relative title length (ln)", 
           Category %in% c("h", "p", "s")) %>%  
    group_by(Category) %>%  
    summarise(y = median(Value)) %>%
    mutate(
      label = as.character(Category)
    ),
                  aes(c(0.75, 1, 1.25), y, label = label),
                  bg.colour = "white", 
    colour = color,
                  bg.r = .2, 
                  force = 0) +
  scale_y_continuous(limits = c(-4.5, 3),
                     breaks =  seq(-5, 5, 1),
                     labels = seq(-5, 5, 1)) +
  scale_x_continuous(labels = NULL) +
  theme_minimal(base_size = base_size) +
  theme(panel.grid  = element_blank()) +
  ylab("Ratio of title length to article length (ln)") +
  xlab("Relative title length")

# Number of pages ------------------
boxlplot_n_pages <- 
items_df %>% 
  ggplot(aes(1,
             log(pages_n))) +
  geom_boxplot(
               size = 1)  +
  # boxplot of data from Fanelli & Glänzel Fig 2
  geom_boxplot(data = sim_data %>% 
  filter(Variable == "N. of pages (ln)",
         Category %in% c("h", "p", "s")),
  aes(1, Value, 
      group = Category),
  size = 1,
  colour = color, 
  alpha = 0)  +
  # annotations from Fanelli & Glänzel Fig 2
  geom_text_repel(data = sim_data %>%  
    filter(Variable == "N. of pages (ln)", 
           Category %in% c("h", "p", "s")) %>%  
    group_by(Category) %>%  
    summarise(y = median(Value)) %>%
    mutate(
      label = as.character(Category)
    ),
                  aes(c(0.75, 1, 1.25), y, label = label),
                  bg.colour = "white",
    colour = color, 
                  bg.r = .2, 
                  force = 0) +
  scale_y_reverse(limits = c(5, 0)) +
  scale_x_continuous(labels = NULL) +
  theme_minimal(base_size = base_size) + 
  theme(panel.grid  = element_blank()) +
  ylab("N. of pages (ln)") +
  xlab("Article length")

@@

# plot
boxlplot_price_index <- 
items_df %>% 
  ggplot(aes(1,
             prices_index)) +
  geom_boxplot(
               size = 1)  +
  # boxplot of data from Fanelli & Glänzel Fig 2
  geom_boxplot(data = sim_data %>% 
  filter(Variable == "Price's index",
         Category %in% c("h", "p", "s")),
  aes(1, Value, 
      group = Category),
  size = 1,
  colour = color,  
  alpha = 0)  +
  # annotations from Fanelli & Glänzel Fig 2
  geom_text_repel(data = sim_data %>%  
    filter(Variable == "Price's index", 
           Category %in% c("h", "p", "s")) %>%  
    group_by(Category) %>%  
    summarise(y = median(Value)) %>%
    mutate(
      label = as.character(Category)
    ),
                  aes(c(0.75, 1, 1.25), y, label = label),
                  bg.colour = "white", colour = color, 
                  bg.r = .2, 
                  force = 0) +
  scale_y_continuous(limits = c(0, 1)) +
  scale_x_continuous(labels = NULL) +
  theme_minimal(base_size = base_size) +
  theme(panel.grid  = element_blank()) +
  ylab("Prop. refs in last 5 years") +
  xlab("Recency of references")

@@

# plot
boxlplot_shannon_index <- 
shannon_per_item %>% 
  filter(!is.na(year)) %>% 
  ggplot(aes(1,
             shannon)) +
  geom_boxplot(
               size = 1)  +
   # boxplot of data from Fanelli & Glänzel Fig 2
  geom_boxplot(data = sim_data %>% 
  filter(Variable == "Shannon div. of sources",
         Category %in% c("h", "p", "s")),
  aes(1, Value, 
      group = Category),
  size = 1,
  colour = color,  
  alpha = 0)  +
  # annotations from Fanelli & Glänzel Fig 2
  geom_text_repel(data = sim_data %>%  
    filter(Variable == "Shannon div. of sources", 
           Category %in% c("h", "p", "s")) %>%  
    group_by(Category) %>%  
    summarise(y = median(Value)) %>%
    mutate(
      label = as.character(Category)
    ),
                  aes(c(0.75, 1, 1.25), y, label = label),
    colour = color, 
                  bg.colour = "white", 
                  bg.r = .2, 
                  force = 0) +
  scale_y_reverse(limits = c(6, 0)) +
  scale_x_continuous(labels = NULL) +
  theme_minimal(base_size = base_size)  +
  theme(panel.grid  = element_blank()) +
  ylab("Shannon Index") +
  xlab("Diversity of references")
```

```{r}
#| label: fig-compare-other-fields
#| fig-cap: Distributions of article characteristics hypothesised to reflect the level of consensus. The boxplot shows the distribution of values of archaeology articles. The thick black line in the middle of the boxplot is the median value, the box represents the inter-quartile range (the range between the 25th and 75th percentiles, where 50% of the data are located), and individual points represent outliers. The smaller coloured boxplots indicate the values computed by Fanelli and Glanzel (2013), where p = physics, s = social sciences, h = humanities. ln denotes the natural logarithm, or logarithm to the base e. 
#| fig-height: 4

@@

@fig-compare-other-fields shows the distribution of bibliometric variables for archaeology in context of data from other fields presented by @fanelliBibliometricEvidenceHierarchy2013. The most striking indicator of archaeology as a hard science is the number of authors, where it is between the social sciences and physics. Archaeology is a close fit with the social sciences in relative title length. It is between the social sciences and humanities in recency of references and diversity of references. The clearest indicator of archaeology as a soft science is article length where it is similar to the humanities. Overall archaeology does not sit squarely at either end of the hard-soft spectrum. It is generally not a harder science than the social sciences, with the exception of collaborator group sizes.

@@

The most striking change over time is in the increase in the number of authors, which has the highest r-squared value of these metrics. One interesting detail evident in @fig-change-over-time is the increase in the range of diversity of references after about 2010. This may be due to some broader changes in academic publishing around this time, such as moves to digital-first continuous publishing, new journals appearing (e.g. *Archaeological and Anthropological Sciences* in 2009 and *Journal of Island & Coastal Archaeology* in 2010), and non-archaeology journals becoming more relevant to archaeologists. For example, *PLOS ONE* received its first impact factor in 2010 and in 2011 Nature's *Scientific Reports* began publishing [@malashichev2017open]. The appearance of Google Scholar in 2004, increasing the discoverability of many works for many researchers, may have also contributed to this increase in range and diversity of references.

@@

The PCA results in panel G show of @fig-variation-by-journal that PC1 captures most of the variance in the metrics (`r variance_explained[1]`%) and is reasonable proxy for the hard-soft spectrum, with *Journal of Cultural Heritage* representing the hard extreme on the right and *Journal of Archaeological Research* representing the soft extreme on the left. The distribution of PC1 values is skewed left, with most of the journals concentrated at the harder end of the spectrum. The distribution of PC2 values reveals additional structure to the data and can be roughly separated into generalist journals in the negative range (e.g. *American Antiquity*, *Antiquity*, *Advances in Archaeological Practice*), and more specialised journals in the positive range (e.g. *Environmental Archaeology*, *Geoarchaeology*, *Archaeologial Research in Asia*, *Journal of Island and Coastal Archaeology*). The *Journal of Archaeological Science* sits about midway between these two groups, reflecting its relevance to specialised communities of practice in archaeology, as well as generalist communities of practice.

@@

With many disciplines making increasing use of computationally intensive analyses in recent years there has been renewed interest in reproducibility [@levequeReproducibleResearchScientific2012]. In part, this is because computationally intensive research is difficult to communicate within the constraints of the methods section of a traditional journal article — the reader also needs the computer code written by the original authors, not just the article text. There is also the broader context of rising pressure to publish in prestigious journals and intense competition for funds that create strong incentives for malpractice in research [@edwards2017academic]. These two factors have led to widespread concerns of a reproducibility crisis in many fields [@Baker2016]. Estimates of scientific reproducibility in several fields confirm the extent of this problem. Empirical replications of 100 studies published in three psychology journals found that 36% of replications had statistically significant results, compared to 96% of the original studies [@OpenScienceCollaboration2015]. Similar empirical replications of large numbers of social science studies and experiemental economics studies successfully replicated 61% and 62% of their target studies respectively [@camererEvaluatingReplicabilitySocial2018; @camererEvaluatingReplicabilityLaboratory2016].

@@

base_size <- 6

@@

```{r}
#| label: fig-aer-summary
#| fig-cap: "Summary of reproducibility reviews for JAS. A: Primary software used for the computational analysis reported in a manuscript. B: Computational or statistical method used by the authors (GMM = geometric morphometrics; Frequentist = hypothesis tests such as chi-square and ANOVA; AI/ML = artifical intellegence and machine learning, including neural networks and deep learning; MCMC = Markov Chain Monte Carlo,  i.e. Bayesian models and other simulations; Network = statistical analysis of social networks; 3D = analysis of 3D data such as artefact models; Composition = compositional analysis of artefacts). C: Locations where authors deposited their code and data files. D: Issues that prevented the reproducible review from succeeding on the first attempt. E: Relationship between software used and issues that make research irreproducible."
#| fig-height: 3

@@

At the time of writing (January 2025) we have completed 47 reproducibility reviews of 25 manuscripts submitted to JAS (most papers required multiple reviews). Of these, 11 have been published in JAS to date. Seven of these eleven papers fully passed the reproducibility review, resulting in a success rate, by one measure, of 63%. Four of the seven papers could be fully reproduced on my first attempt, the others required additional input from the authors. For comparison with reproducibility studies in other fields reported above, the seven fully reproducible papers should be divided by the 25 reviewed for reproducibility, resulting in a 28% success rate. Expanding the denominator to include the total number of research articles published in JAS from May 2024 (when the first article to pass the reproducibility review, @herskindComputationalLinguisticMethodology2024, was published) to January 2025 (n = 97), which are articles that could have been eligible for reproducibility review had the authors used an open source programming language (e.g. instead of commercial software such as Microsoft Excel or SPSS, etc.) the success rate is 7%. In any case, the computational reproducibility of archaeological research is generally on the low end of the distribution of values available from a variety of hard and soft sciences.

@@

Third, authors should include in their compendium a README document that describes to readers the folders and files contained in the project. The README file is typically the first file that a reader will look at in a compendium so it should include brief instructions to guide the user to a successful reproduction of the original results (e.g. what order to run the code files in). A README should also briefly describe the contents of the compendium, where other necessary files can be obtained (e.g. data files not included in the compendium), the key software packages needed and the version numbers that the authors used, and if the analysis takes more than a few minutes to run on a typical laptop, the hardware resources and compute time used by the author.

@@

Computational reproducibility is not a panacea, results that are fully reproducible can contain errors and fraud. It is no guarantee of code quality, or that statistics have been used appropriately (cf. @Vaiglova_2025; Crema 2025), or that data management is consistent with FAIR and CARE principles [@carrollOperationalizingCAREFAIR2021]. It is also time-consuming for authors to ensure their computational work can be reproduced, and for reviewers to evaluate. In a professional environment where job security and career progression is often associated with pressure to publish many high-impact papers, demands for authors to spend time on reproducibility, resulting in less time for publishing more papers, may seem frustrating. This may seem especially unfair to early career researchers on short-term contracts, who may feel the goalposts for career success are being moved and that they are being asked to do more work their graduate training has not prepared them for. This highlights the need for a culture shift among senior archaeological scientists to value reproducibility in hiring and promotion decision-making. Professors must contribute to this shift by nurturing a culture of reanalysis and reproducibility in their teaching, for example by using replication assignments and by training students in the most current best practices and tools for reproducible research, such as R and Python [@marwickHowUseReplication2020].
